Should geophysicists use the gravity disturbance or the anomaly?

### 2018-06-18 manuscript/letter.tex
change the latex template

% gjilguid2e.tex
% V2.0 released 1998 December 18
% V2.1 released 2003 October 7 -- Gregor Hutton, updated the web address for the style files.

%\documentclass[extra,mreferee]{gji}
\documentclass[extra]{gji}
\usepackage{timet}

%Extra packages
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage[pdftex,colorlinks=true]{hyperref}
\hypersetup{
    allcolors=black,
}

\title[Gravity anomaly or gravity disturbance?]
      {Should geophysicists use the gravity disturbance or the anomaly?}

\author[Oliveira Jr et al.]{
\parbox{\linewidth}{%
    Vanderlei C. Oliveira Jr$^1$\thanks{Email: vanderlei@on.br},
    Leonardo Uieda$^2$, Kristoffer A. T. Hallam$^{1}$ and
    \linebreak Val\'{e}ria C. F. Barbosa$^1$
    \vspace{0.3cm} }%
    \\
    $^1$ Department of Geophysics, Observat\'{o}rio Nacional, Rio de Janeiro, Brazil \\
    $^2$ Department of Geology and Geophysics, SOEST, University of Hawai'i at M\={a}noa, Honolulu, USA
}

\date{Received 2018 Month XX; in original form 2018 Month XX}
\pagerange{\pageref{firstpage}--\pageref{lastpage}}
\volume{XXX}
\pubyear{2018}

%\def\LaTeX{L\kern-.36em\raise.3ex\hbox{{\small A}}\kern-.15em
%    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}
%\def\LATeX{L\kern-.36em\raise.3ex\hbox{{\Large A}}\kern-.15em
%    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}
% Authors with AMS fonts and mssymb.tex can comment out the following
% line to get the correct symbol for Geophysical Journal International.
\let\leqslant=\leq

\newtheorem{theorem}{Theorem}[section]

\newcommand{\versor}[1]{\hat{\mitbf{#1}}}
\renewcommand{\vector}[1]{\vec{\mitbf{#1}}}

@@

\label{firstpage}

@@

\begin{summary}
 The gravity anomaly is defined as the difference between the Earth's gravity
 on the geoid and the normal gravity on the reference ellipsoid.
 Because these quantities are not at the same point, the anomaly contains
 centrifugal accelerations and cannot be considered a harmonic function.
 The gravity disturbance is the difference between gravity and normal gravity
 at the same point.
 Consequently, the centrifugal effects can be neglected and the disturbance can
 be considered a harmonic function.
 This is the premise behind most potential-field data processing techniques
 (e.g., upward/downward continuation).
 Unlike the anomaly, the disturbance is due solely to the
 gravitational effects of geologic sources, making it the most appropriate
 for geophysical purposes.
 Use of the gravity anomaly in geophysics carries with it the implicit
 assumption that it is a good approximation for the gravity disturbance.
 However, bear in mind that the difference between the gravity disturbance and
 the free-air anomaly can be larger than 10 mGal worldwide.
 In fact, we argue that the assumptions made during gravity forward and inverse
 modeling imply that the quantity being modelled is the disturbance, not the
 anomaly.
\end{summary}

\begin{keywords}
 Geopotential theory -- Gravity anomalies and Earth structure -- Reference systems -- Numerical modelling.
\end{keywords}

@@

The first step toward isolating the gravitational component
is to remove the effects of vehicle motion, Earth tides, instrumental drift,
and barometric pressure changes, among others.
If these effects are properly removed,
the resultant observations are considered to be solely
the sum of the centrifugal acceleration due to the Earth's rotation and
the gravitational acceleration produced by
the internal density distribution of the whole Earth.
The isolation of this particular gravitational component,
and its subsequent use for estimating density
distributions related to geological structures in subsurface,
are the main goals of applied gravimetry \citep{blakely1996}.
The computation of the gravitational effect produced by
the geological structures is known in geophysics as
\textit{gravity modelling}.
Not to be confused with \textit{gravity field modelling}, which
is commonly used in geodesy to denote the mathematical characterization of the
gravity field in local, regional, or global scales (e.g., spherical harmonic
analysis).

We present a discussion of whether the gravity disturbance or the gravity
anomaly should be used in geophysical applications.
The concepts discussed here are well-established in the literature.
However, the debate around this theoretical issue has been
carried about from a more geodetic than a geophysical point of view
\citep{lafehr1991,chapin1996,li2001,fairhead2003,
hackney-featherstone2003,hinze2005}.
We will approach the topic by comparing the geophysical interpretations of the
disturbance and the anomaly.
Our reasoning suggests that the gravity disturbance is more appropriated for
geophysical purposes than the gravity anomaly.

@@

\section{Terrestrial Reference Systems used in gravity modelling}

For geophysical gravity modelling purposes, there are three important
Terrestrial Reference Systems.
They rotate with the Earth and are used for describing
positions and movements of objects on and close to the Earth's surface
\citep{torge2012}.

The first is a geocentric system of Cartesian coordinates
having the $Z$-axis coincident with the mean rotational axis of the Earth,
the $X$-axis pointing to the Greenwich meridian,
and the $Y$-axis completing a right-handed system (Fig. \ref{fig:systems}).
This reference system is called by different names in the literature:
Mean Terrestrial System \citep[e.g.,][]{soler1976},
Earth-fixed geocentric Cartesian system \citep[e.g.,][]{torge2012}
or Earth-centered Earth-fixed system \citep[e.g.,][]{bouman_etal2013},
for example.
Here, we opted for using the term Geocentric Cartesian System (GCS).

The second is a geocentric system of geodetic coordinates
(the Geocentric Geodetic System or GGS),
which is defined by the reference ellipsoid used as the Normal Earth
\citep{heiskanen-moritz1967, soler1976, torge2012, bouman_etal2013}.
The position of a point is defined by
the \textit{geometric height} $h$,
the \textit{geodetic latitude} $\varphi$,
and the \textit{longitude} $\lambda$ (Fig. \ref{fig:systems}).
There are also three mutually-orthogonal unit vectors defined at a given point
\citep{soler1976}:

\begin{equation}
    \versor{u} =
    \begin{bmatrix}
        \cos\varphi \, \cos\lambda \\
        \cos\varphi \, \sin\lambda \\
        \sin\varphi
    \end{bmatrix}
    \versor{v} =
        \begin{bmatrix}
        -\sin\varphi \, \cos\lambda \\
        -\sin\varphi \, \sin\lambda \\
        \cos\varphi
    \end{bmatrix}
    \versor{w} =
    \begin{bmatrix}
        -\sin\lambda \\
        \cos\lambda \\
        0
    \end{bmatrix} .
    \label{eq:unit-vectors}
\end{equation}

@@

The third is the Topocentric Cartesian System (TCS),
which is commonly used in local or regional scale geophysical studies.
The origin of the TCS is located at a point $P$,
axes $x$ and $y$ are parallel to
the unit vectors $\versor{v}$ and $\versor{w}$
(eq. \ref{eq:unit-vectors}), respectively,
and the $z$-axis is opposite to the unit vector $\versor{u}$
(Fig. \ref{fig:systems}c).

\begin{figure}
%    \vspace{5.5cm}
    \includegraphics{figures/cartesian-geodetic-systems.eps}
    \caption{
    (a) and (b) The Geocentric Cartesian System (GCS) and the Geocentric
    Geodetic System (GGS).
    The GGS is defined by an oblate ellipsoid with semi-minor axis $b$ and
    semi-major axis $a$.
    In this coordinate system, the position of a point is determined by the
    geometric height $h$, geodetic latitude $\varphi$, and longitude $\lambda$.
    $O$ is the Earth's center of mass, $P$ is an observation point, and
    unit vectors $\versor{u}$, $\versor{v}$, and $\versor{w}$ define mutually
    orthogonal directions at $P$ (eq. \ref{eq:unit-vectors}).
    In (b), $Q$ is the projection of $P$ onto the reference ellipsoid at
    the same latitude and longitude.
    (c) The topocentric Cartesian coordinate system (TCS) with origin at a
    point $P$.
    Axes $x$ and $y$ are parallel to the unit vectors $\versor{v}$ and
    $\versor{w}$, respectively, and the $z$ axis is opposite to $\versor{u}$.
    The gray plane is the same shown in (a) and (b).
    }
  \label{fig:systems}
\end{figure}

\begin{figure}
    \includegraphics{figures/surfaces.eps}
    \caption{
    The gravity vector
    $\vector{g}_{P}$, normal gravity vector $\vector{\gamma}_{P}$,
    gravity disturbance vector $\delta \vector{g}_{P}$
    (eq. \ref{eq:gravity-disturbance-vector}), and unit vector
    $\versor{u}_{P}$ (eq. \ref{eq:unit-vectors}) at a point $P$
    on the surface of the Earth.
    The gravity vector $\vector{g}_{Q^{\prime}}$
    at a point $Q^{\prime}$ on the geoid, normal gravity vector
    $\vector{\gamma}_{Q}$ at a point $Q$ on the reference ellipsoid,
    geometric height $h$, orthometric height $H$, and geoidal height $N$.
    The dashed line passing through
    $Q$, $Q^{\prime}$ and $P$ is normal to the
    surface of the reference ellipsoid at $Q$.
    This figure shows a commonly used approximation in which the ellipsoid and
    geoid are represented as parallel surfaces so that $h \approx H + N$.
    }
  \label{fig:surfaces}
\end{figure}

@@

Let $\vector{\gamma}_{P}$ and $\vector{g}_{P}$ and be, respectively,
the \textit{normal gravity vector} and the Earth's \textit{gravity vector}
(corrected from non-gravitational effects due to vehicle
motion and time variations such as Earth tides and instrumental drift),
both located at a point $P$.
In this case, the gravity vector $\vector{g}_{P}$ is the
gradient of the scalar \textit{gravity potential},
which is the sum of a gravitational potential and a centrifugal potential.
Similarly, the normal gravity vector $\vector{\gamma}_{P}$ is the
gradient of the scalar \textit{normal potential},
which is also the sum of a gravitational and a centrifugal potential.
By definition, the centrifugal part of
the normal potential and the gravity potential are equal.

The difference between $\vector{g}_{P}$ and $\vector{\gamma}_{P}$,
which we emphasize are located at the same point $P$,
is the \textit{gravity disturbance vector} (Fig. \ref{fig:surfaces}):

\begin{equation}
\delta \vector{g}_{P} =
\vector{g}_{P} - \vector{\gamma}_{P} \: .
\label{eq:gravity-disturbance-vector}
\end{equation}

Because the centrifugal parts of both the normal gravity vector
and the Earth's gravity vector are equal,
the gravity disturbance vector $\delta \vector{g}_{P}$
represents a purely gravitational effect.
As a consequence of the superposition principal \citep{blakely1996},
this effect is caused by contrasts between
the internal density distributions
of the actual Earth and the normal Earth.
In applied geophysics, these density differences are generally called
\textit{anomalous masses} \citep[e.g.,][]{hammer1945,lafehr1965},
\textit{density anomalies} \citep[e.g.,][]{forsberg1984},
or \textit{gravity sources} \citep[e.g.,][]{blakely1996}.
Here, we opted for using the last term.

The difference between the magnitudes of the gravity vector
$g_{P} = \| \vector{g}_{P} \|$ and the normal gravity vector
$\gamma_{P} = \| \vector{\gamma}_{P} \|$,
at the same point $P$, is the \textit{gravity disturbance}
\citep{heiskanen-moritz1967, hofmann-wellenhof-moritz2005}:

@@

\noindent
Notice that $\delta g_{P}$ is not equivalent
to the magnitude of the gravity disturbance vector
$\delta \vector{g}_{P}$ \citep{barthelmes2013, sanso_sideris2013}.

The \textit{gravity anomaly vector}
is the difference between the gravity
vector at a point $Q^{\prime}$ on the geoid
(a particular equipotential surface of the gravity potential)
and the normal gravity vector at a point $Q$ on the surface of the ellipsoid,
both located at the same geodetic latitude and longitude
(Fig. \ref{fig:surfaces}):

\begin{equation}
\Delta \vector{g} = \vector{g}_{Q^{\prime}} - \vector{\gamma}_{Q} .
\label{eq:gravity-anomaly-vector}
\end{equation}

Similarly to the gravity disturbance (eq. \ref{eq:gravity-disturbance}),
the \textit{gravity anomaly} is given by

@@

\noindent
where $g_{Q^{\prime}}$ is the magnitude of the gravity vector on the geoid,
and $\gamma_{Q}$ is the magnitude of the normal gravity vector
on the reference ellipsoid (Fig. \ref{fig:surfaces}).

@@

\noindent
where the rightmost term is called the \textit{free-air correction},
$\partial \gamma/\partial h \approx -0.3086$ mGal/m is the
derivative of the normal gravity with respect to the geometric height $h$,
and $H_{P}$ is the orthometric height (i.e., with respect to the geoid)
of point $P$ (Fig. \ref{fig:surfaces}).
The free-air correction can be interpreted as a downward continuation of
$g_P$ to the geoid or as an upward continuation of $\gamma_Q$ to a height
$H_P$.
Regardless, we can also apply the free-air correction
to calculate the gravity disturbance by
approximating the normal gravity $\gamma_{P}$
at a point $P$ on the Earth's surface
from the normal gravity at the ellipsoid $\gamma_{Q}$:

@@

\noindent
where $N \approx h - H$ is the geoidal height (Fig. \ref{fig:surfaces}).
This approximation assumes that the geoid and the surface of
the reference ellipsoid are close to being parallel at $P$
and the surrounding area.
Eqs \ref{eq:gravity-disturbance-approx}
and \ref{eq:difference-disturbance-free-air} are commonly used
in geodesy to define the \textit{fundamental equation of physical
geodesy} \citep{hofmann-wellenhof-moritz2005}.

We know empirically that $N$ is $\approx \pm 1$ m on the oceans
and reaches a maximum absolute value of $\approx 120$ m
\citep[e.g.,][]{torge2012, sanso_sideris2013}.
Fig. \ref{fig:difference} shows the absolute difference between the
gravity disturbance and the free-anomaly globally calculated
using eq. \ref{eq:difference-disturbance-free-air} and the spherical harmonic
model EIGEN-6c4 \citep{forste2014}.
In Brazil, for example, the geoidal height
varies between $\approx \pm 30$ m \citep{ibge_mapgeo2015}.
Consequently,
the maximum absolute difference between the gravity disturbance and
the free-air anomaly in Brazil,
according to eq. \ref{eq:difference-disturbance-free-air},
is $\approx 9.3$ mGal.

\begin{figure}
    \includegraphics[width=\columnwidth]{figures/difference.eps}
    \caption{
    Absolute difference between the gravity disturbance and free-air anomaly.
    Calculated using eq. \ref{eq:difference-disturbance-free-air}, the
    EIGEN-6c4 \citep{forste2014} spherical harmonic model, and software
    \textit{Fatiando a Terra} \citep{uieda2013}. The figure was made using GMT
    \citep{wessel2013}.
    }
  \label{fig:difference}
\end{figure}

@@

\noindent
where $\rho_t$ is the density of the topographic masses and $G$ is the
gravitational constant.
The Bouguer correction approximates the gravitational effect that all
topographic masses above the geoid exert on point $P$ by the effect of a
homogeneous, infinitely extended slab of constant density $\rho_{t}$ and
thickness $H_{P}$.
A similar correction can be applied to the gravity disturbance
(using the approximation in eq. \ref{eq:gravity-disturbance-approx})
to remove the effect of topographic masses between
the observation point $P$ and a point on the ellipsoid $Q$.
We will call this quantity the \textit{Bouguer disturbance}:

@@

At first glance, the only difference between eqs.
\ref{eq:simple-bouguer-anomaly} and \ref{eq:bouguer-disturbance} is the use of
orthometric ($H_P$) versus geometric ($h_P$) heights.
However, their geophysical interpretations are significantly different.
Apart from the centrifugal effect mentioned in the previous section,
the Bouguer anomaly (eq. \ref{eq:simple-bouguer-anomaly}) still contains
the gravitational effect of topographic masses between the geoid and ellipsoid.
Conversely, when calculating the Bouguer disturbance one removes the effects of
an ellipsoidal reference model (the normal Earth) and the known topographic
masses between the surface of the ellipsoid and observation point.

@@

\section{Gravity modelling}

We have established that the gravity disturbance is due solely to the
gravitational effect of the gravity sources.
Now we will discuss how this relates to some assumptions made during gravity
modelling.

The gravity vector $\vector{g}_i$ (eq. \ref{eq:gravity-disturbance-vector})
at a point $(x_i, y_i, z_i)$ in the TCS (Fig. \ref{fig:systems}c) can be
represented by:

\begin{equation}
\vector{g}_i = \vector{\gamma}_i + \delta \vector{g}_i \: .
\label{eq:gravity-vector-TCS}
\end{equation}

@@

\begin{equation}
g_i \approx \gamma_i +
\versor{\gamma}_i^{\top} \delta \vector{g}_i \: ,
\label{eq:gobs-approx}
\end{equation}

\noindent
where $^{\top}$ denotes transposition and $\versor{\gamma}_{i}$ is a unit
vector with the direction of the normal gravity vector.
This approximation is valid because
$\gamma_{i} \gg \| \delta \mathbf{g}_{i} \|$ at all points on or above the
Earth's surface.
It is known in geodesy \citep[e.g.,][]{sanso_sideris2013}
and is analogous to the approximation used for
total-field magnetic anomalies \citep[e.g.,][]{blakely1996}.
In local or regional scales, the unit
vector $\versor{\gamma}_{i}$
may be considered constant throughout the study area and
parallel to the $z$ axis of the TCS (Fig. \ref{fig:systems}c).

Using eq. \ref{eq:gobs-approx} and fixing $\gamma$ at a point $P$ at
the origin of the TCS (Fig. \ref{fig:systems}c),
the gravity disturbance can be rewritten as:

\begin{equation}
\delta g_{i} \approx \versor{\gamma}_{P}^{\top} \delta \vector{g}_{i} \: .
\label{eq:gravity-disturbance-approx-TCS}
\end{equation}

\noindent
This equation shows that the gravity disturbance $\delta g_{i}$ is
approximately the component of the $\delta \vector{g}_{i}$ in the direction of
the normal gravity vector $\vector{\gamma}_P$
\citep{hofmann-wellenhof-moritz2005, sanso_sideris2013}.
Because $z$-axis of the TCS is parallel to $\vector{\gamma}_P$
(Fig. \ref{fig:surfaces}), the gravity disturbance can be defined as the
$z$-component of the gravitational attraction exerted by the gravity sources at
the point $(x_{i}, y_{i}, z_{i})$.
As a consequence, the gravity disturbance can be modelled by the harmonic
function:

\begin{equation}
d_{i} = \iiint\limits_{v}
\frac{G \, \Delta\rho(x^{\prime}, y^{\prime}, z^{\prime}) \,
[z_{i} - z^{\prime}]}
{\sqrt{[x_{i} - x^{\prime}]^{2} +
[y_{i} - y^{\prime}]^{2} + [z_{i} - z^{\prime}]^{2}}} \: dv^{\prime} \: ,
\label{eq:gz-local}
\end{equation}

@@

In practice, most instances of gravity modelling in the literature use Bouguer
anomalies and function $d_{i}$ (usually presented as ``$g_z$'') to represent
the gravitational effect of gravity sources.
Consequently, these studies are using the gravity
anomaly as an approximation for the gravity disturbance.
Notice that $d_{i}$ does not depend on the
height with respect to the geoid (orthometric height).
Rather, it depends on the relative position of the gravity sources and the
observation point $(x_{i}, y_{i}, z_{i})$ in the TCS (Fig. \ref{fig:systems}c).
Finally, it is important to mention that gravimeters do not measure $d_{i}$ but
$\| \vector{g}_i \|$.

>>>>

\documentclass[paper,twocolumn,twoside]{geophysics}
%\documentclass[manuscript,revised]{geophysics}

% An example of defining macros
\newcommand{\rs}[1]{\mathstrut\mbox{\scriptsize\rm #1}}
\newcommand{\rr}[1]{\mbox{\rm #1}}

% Extra packages
\usepackage{amsmath}
\usepackage{bm}
\usepackage[pdftex,colorlinks=true]{hyperref}
\hypersetup{
	allcolors=black,
}

\renewcommand{\figdir}{figures} % figure directory

@@

\title{Should geophysicists use the gravity disturbance or the anomaly?}

\renewcommand{\thefootnote}{\fnsymbol{footnote}}

\ms{GEO-XXXX} % manuscript number

\address{
\footnotemark[1]Observat\'{o}rio Nacional, \\
Department of Geophysics, \\
Rio de Janeiro, Brazil \\
\footnotemark[2]University of Hawai'i at M\={a}noa, \\
Department of Geology and Geophysics, SOEST, \\
Honolulu, USA}
\author{Vanderlei C. Oliveira Jr\footnotemark[1], Leonardo Uieda\footnotemark[2],
Kristoffer A. T. Hallam\footnotemark[1] and Val\'{e}ria C. F. Barbosa\footnotemark[1]}

\footer{vanderlei@on.br, leouieda@gmail.com, kristoffer.hallam@gmail.com, valcris@on.br}
\lefthead{Oliveira Jr et al.}
\righthead{Gravity anomaly or gravity disturbance?}

@@

\begin{abstract}
  The gravity anomaly is defined as the difference between the Earth's gravity
 on the geoid and the normal gravity on the reference ellipsoid.
 Because these quantities are not at the same point, the anomaly contains
 centrifugal accelerations and cannot be considered a harmonic function.
 The gravity disturbance is the difference between gravity and normal gravity
 at the same point.
 Consequently, the centrifugal effects can be neglected and the disturbance can
 be considered a harmonic function.
 This is the premise behind most potential-field data processing techniques
 (e.g., upward/downward continuation).
 Unlike the anomaly, the disturbance is due solely to the
 gravitational effects of geologic sources, making it the most appropriate
 for geophysical purposes.
 Use of the gravity anomaly in geophysics carries with it the implicit
 assumption that it is a good approximation for the gravity disturbance.
 However, bear in mind that the difference between the gravity disturbance and
 the free-air anomaly can be larger than 10 mGal worldwide.
 In fact, we argue that the assumptions made during gravity forward and inverse
 modeling imply that the quantity being modeled is the disturbance, not the
 anomaly.
\end{abstract}

@@

The first step toward isolating the gravitational component
is to remove the effects of vehicle motion, Earth tides, instrumental drift,
and barometric pressure changes, among others.
If these effects are properly removed,
the resultant observations are considered to be solely
the sum of the centrifugal acceleration due to the Earth's rotation and
the gravitational acceleration produced by
the internal density distribution of the whole Earth.
The isolation of this particular gravitational component,
and its subsequent use for estimating density
distributions related to geological structures in subsurface,
are the main goals of applied gravimetry \citep{blakely1996}.
The computation of the gravitational effect produced by
the geological structures is known in geophysics as
\textit{gravity modeling}.
Not to be confused with \textit{gravity field modeling}, which
is commonly used in geodesy to denote the mathematical characterization of the
gravity field in local, regional, or global scales (e.g., spherical harmonic
analysis).

We present a discussion of whether the gravity disturbance or the gravity
anomaly should be used in geophysical applications.
The concepts discussed here are well-established in the literature and
are familiar to most geodesists.
However, the debate around this theoretical issue has been
carried about from a more geodetic than a geophysical point of view
\citep{lafehr1991,chapin1996,li2001,fairhead2003,
hackney-featherstone2003,hinze2005} and maybe that is why it is
unfamiliar to most exploration geophysicists.
We will approach the topic by comparing the geophysical interpretations of the
disturbance and the anomaly.
Our reasoning suggests that the gravity disturbance is more appropriated for
geophysical purposes than the gravity anomaly.

@@

\section{Terrestrial Reference Systems used in gravity modeling}

For geophysical gravity modeling purposes, there are three important
Terrestrial Reference Systems.
They rotate with the Earth and are used for describing
positions and movements of objects on and close to the Earth's surface
\citep{torge2012}.

The first is a geocentric system of Cartesian coordinates
having the $Z$-axis coincident with the mean rotational axis of the Earth,
the $X$-axis pointing to the Greenwich meridian,
and the $Y$-axis completing a right-handed system (Figure \ref{fig:systems}).
This reference system is called by different names in the literature:
Mean Terrestrial System \citep[e.g.,][]{soler1976},
Earth-fixed geocentric Cartesian system \citep[e.g.,][]{torge2012}
or Earth-centered Earth-fixed system \citep[e.g.,][]{bouman_etal2013},
for example.
Here, we opted for using the term Geocentric Cartesian System (GCS).

The second is a geocentric system of geodetic coordinates
(the Geocentric Geodetic System or GGS),
which is defined by the reference ellipsoid used as the Normal Earth
\citep{heiskanen-moritz1967, soler1976, torge2012, bouman_etal2013}.
The position of a point is defined by
the \textit{geometric height} $h$,
the \textit{geodetic latitude} $\varphi$,
and the \textit{longitude} $\lambda$ (Figure \ref{fig:systems}).
There are also three mutually-orthogonal unit vectors defined at a given point
\citep{soler1976}:

\begin{equation}
    \mathbf{u} =
    \begin{pmatrix}
        \cos\varphi \, \cos\lambda \\
        \cos\varphi \, \sin\lambda \\
        \sin\varphi
    \end{pmatrix}
    \mathbf{v} =
    \begin{pmatrix}
        -\sin\varphi \, \cos\lambda \\
        -\sin\varphi \, \sin\lambda \\
        \cos\varphi
    \end{pmatrix}
    \mathbf{w} =
    \begin{pmatrix}
        -\sin\lambda \\
        \cos\lambda \\
        0
    \end{pmatrix} \: .
    \label{eq:unit-vectors}
\end{equation}

@@

The third is the Topocentric Cartesian System (TCS),
which is commonly used in local or regional scale geophysical studies.
The origin of the TCS is located at a point $P$,
axes $x$ and $y$ are parallel to
the unit vectors $\mathbf{v}$ and $\mathbf{w}$
(equation \ref{eq:unit-vectors}), respectively,
and the $z$-axis is opposite to the unit vector $\mathbf{u}$
(Figure \ref{fig:systems}c).

@@

Let $\boldsymbol{\gamma}_{P}$ and $\mathbf{g}_{P}$ and be, respectively,
the \textit{normal gravity vector} and the Earth's \textit{gravity vector}
(corrected from non-gravitational effects due to vehicle
motion and time variations such as Earth tides and instrumental drift),
both located at a point $P$.
In this case, the gravity vector $\mathbf{g}_{P}$ is the
gradient of the scalar \textit{gravity potential},
which is the sum of a gravitational potential and a centrifugal potential.
Similarly, the normal gravity vector $\boldsymbol{\gamma}_{P}$ is the
gradient of the scalar \textit{normal potential},
which is also the sum of a gravitational and a centrifugal potential.
By definition, the centrifugal part of
the normal potential and the gravity potential are equal.

The difference between $\mathbf{g}_{P}$ and $\boldsymbol{\gamma}_{P}$,
which we emphasize are located at the same point $P$,
is the \textit{gravity disturbance vector} (Figure \ref{fig:surfaces}):

\begin{equation}
\delta \mathbf{g}_{P} =
\mathbf{g}_{P} - \boldsymbol{\gamma}_{P} \: .
\label{eq:gravity-disturbance-vector}
\end{equation}

Because the centrifugal parts of both the normal gravity vector
and the Earth's gravity vector are equal,
the gravity disturbance vector $\delta \mathbf{g}_{P}$
represents a purely gravitational effect.
As a consequence of the superposition principal \citep{blakely1996},
this effect is caused by contrasts between
the internal density distributions
of the actual Earth and the normal Earth.
In applied geophysics, these density differences are generally called
\textit{anomalous masses} \citep[e.g.,][]{hammer1945,lafehr1965},
\textit{density anomalies} \citep[e.g.,][]{forsberg1984},
or \textit{gravity sources} \citep[e.g.,][]{blakely1996}.
Here, we opted for using the last term.

The difference between the magnitudes of the gravity vector
$g_{P} = \| \mathbf{g}_{P} \|$ and the normal gravity vector
$\gamma_{P} = \| \boldsymbol{\gamma}_{P} \|$,
at the same point $P$, is the \textit{gravity disturbance}
\citep{heiskanen-moritz1967, hofmann-wellenhof-moritz2005}:

@@

\noindent
Notice that $\delta g_{P}$ is not equivalent
to the magnitude of the gravity disturbance vector
$\delta \mathbf{g}_{P}$ \citep{barthelmes2013, sanso_sideris2013}.

The \textit{gravity anomaly vector}
is the difference between the gravity
vector at a point $Q^{\prime}$ on the geoid
(a particular equipotential surface of the gravity potential)
and the normal gravity vector at a point $Q$ on the surface of the ellipsoid,
both located at the same geodetic latitude and longitude
(Figure \ref{fig:surfaces}):

\begin{equation}
\Delta \mathbf{g} = \mathbf{g}_{Q^{\prime}} - \boldsymbol{\gamma}_{Q} .
\label{eq:gravity-anomaly-vector}
\end{equation}

Similarly to the gravity disturbance (equation \ref{eq:gravity-disturbance}),
the \textit{gravity anomaly} is given by

@@

\noindent
where $g_{Q^{\prime}}$ is the magnitude of the gravity vector on the geoid,
and $\gamma_{Q}$ is the magnitude of the normal gravity vector
on the reference ellipsoid (Figure \ref{fig:surfaces}).

@@

\noindent
where the rightmost term is called the \textit{free-air correction},
$\partial \gamma/\partial h \approx -0.3086$ mGal/m is the
derivative of the normal gravity with respect to the geometric height $h$,
and $H_{P}$ is the orthometric height (i.e., with respect to the geoid)
of point $P$ (Figure \ref{fig:surfaces}).
The free-air correction can be interpreted as a downward continuation of
$g_P$ to the geoid or as an upward continuation of $\gamma_Q$ to a height
$H_P$.
Regardless, we can also apply the free-air correction
to calculate the gravity disturbance by
approximating the normal gravity $\gamma_{P}$
at a point $P$ on the Earth's surface
from the normal gravity at the ellipsoid $\gamma_{Q}$:

@@

\noindent
where $N \approx h - H$ is the geoidal height (Figure \ref{fig:surfaces}).
This approximation assumes that the geoid and the surface of
the reference ellipsoid are close to being parallel at $P$
and the surrounding area.
Eqs \ref{eq:gravity-disturbance-approx}
and \ref{eq:difference-disturbance-free-air} are commonly used
in geodesy to define the \textit{fundamental equation of physical
	geodesy} \citep{hofmann-wellenhof-moritz2005}.

We know empirically that $N$ is $\approx \pm 1$ m on the oceans
and reaches a maximum absolute value of $\approx 120$ m
\citep[e.g.,][]{torge2012, sanso_sideris2013}.
Figure \ref{fig:difference} shows the absolute difference between the
gravity disturbance and the free-anomaly globally calculated
using equation \ref{eq:difference-disturbance-free-air} and the spherical harmonic
model EIGEN-6c4 \citep{forste2014}.
In Brazil, for example, the geoidal height
varies between $\approx \pm 30$ m \citep{ibge_mapgeo2015}.
Consequently,
the maximum absolute difference between the gravity disturbance and
the free-air anomaly in Brazil,
according to equation \ref{eq:difference-disturbance-free-air},
is $\approx 9.3$ mGal.

@@

\noindent
where $\rho_t$ is the density of the topographic masses and $G$ is the
gravitational constant.
The Bouguer correction approximates the gravitational effect that all
topographic masses above the geoid exert on point $P$ by the effect of a
homogeneous, infinitely extended slab of constant density $\rho_{t}$ and
thickness $H_{P}$.
A similar correction can be applied to the gravity disturbance
(using the approximation in equation \ref{eq:gravity-disturbance-approx})
to remove the effect of topographic masses between
the observation point $P$ and a point on the ellipsoid $Q$.
We will call this quantity the \textit{Bouguer disturbance}:

@@

At first glance, the only difference between eqs.
\ref{eq:simple-bouguer-anomaly} and \ref{eq:bouguer-disturbance} is the use of
orthometric ($H_P$) versus geometric ($h_P$) heights.
However, their geophysical interpretations are significantly different.
Apart from the centrifugal effect mentioned in the previous section,
the Bouguer anomaly (equation \ref{eq:simple-bouguer-anomaly}) still contains
the gravitational effect of topographic masses between the geoid and ellipsoid.
Conversely, when calculating the Bouguer disturbance one removes the effects of
an ellipsoidal reference model (the normal Earth) and the known topographic
masses between the surface of the ellipsoid and observation point.

@@

\section{Gravity modeling}

We have established that the gravity disturbance is due solely to the
gravitational effect of the gravity sources.
Now we will discuss how this relates to some assumptions made during gravity
modeling.

The gravity vector $\mathbf{g}_i$ (equation \ref{eq:gravity-disturbance-vector})
at a point $(x_i, y_i, z_i)$ in the TCS (Figure \ref{fig:systems}c) can be
represented by:

\begin{equation}
\mathbf{g}_i = \mathbf{\gamma}_i + \delta \mathbf{g}_i \: .
\label{eq:gravity-vector-TCS}
\end{equation}

@@

\begin{equation}
g_i \approx \gamma_i +
\mathbf{\gamma}_i^{\top} \delta \mathbf{g}_i \: ,
\label{eq:gobs-approx}
\end{equation}

\noindent
where $^{\top}$ denotes transposition and $\mathbf{\gamma}_{i}$ is a unit
vector with the direction of the normal gravity vector.
This approximation is valid because
$\gamma_{i} \gg \| \delta \mathbf{g}_{i} \|$ at all points on or above the
Earth's surface.
It is known in geodesy \citep[e.g.,][]{sanso_sideris2013}
and is analogous to the approximation used for
total-field magnetic anomalies \citep[e.g.,][]{blakely1996}.
In local or regional scales, the unit
vector $\mathbf{\gamma}_{i}$
may be considered constant throughout the study area and
parallel to the $z$ axis of the TCS (Figure \ref{fig:systems}c).

Using equation \ref{eq:gobs-approx} and fixing $\gamma$ at a point $P$ at
the origin of the TCS (Figure \ref{fig:systems}c),
the gravity disturbance can be rewritten as:

\begin{equation}
\delta g_{i} \approx \mathbf{\gamma}_{P}^{\top} \delta \mathbf{g}_{i} \: .
\label{eq:gravity-disturbance-approx-TCS}
\end{equation}

\noindent
This equation shows that the gravity disturbance $\delta g_{i}$ is
approximately the component of the $\delta \mathbf{g}_{i}$ in the direction of
the normal gravity vector $\mathbf{\gamma}_P$
\citep{hofmann-wellenhof-moritz2005, sanso_sideris2013}.
Because $z$-axis of the TCS is parallel to $\mathbf{\gamma}_P$
(Figure \ref{fig:surfaces}), the gravity disturbance can be defined as the
$z$-component of the gravitational attraction exerted by the gravity sources at
the point $(x_{i}, y_{i}, z_{i})$.
As a consequence, the gravity disturbance can be modelled by the harmonic
function:

\begin{equation}
d_{i} = \iiint\limits_{v}
\frac{(z_{i} - z^{\prime}) \, G \, \Delta\rho(x^{\prime}, y^{\prime}, z^{\prime})}
{\sqrt{(x_{i} - x^{\prime})^{2} +
		(y_{i} - y^{\prime})^{2} + (z_{i} - z^{\prime})^{2}}} \: dv^{\prime} \: ,
\label{eq:gz-local}
\end{equation}

@@

In practice, most instances of gravity modeling in the literature use Bouguer
anomalies and function $d_{i}$ (usually presented as ``$g_z$'') to represent
the gravitational effect of gravity sources.
Consequently, these studies are using the gravity
anomaly as an approximation for the gravity disturbance.
Notice that $d_{i}$ does not depend on the
height with respect to the geoid (orthometric height).
Rather, it depends on the relative position of the gravity sources and the
observation point $(x_{i}, y_{i}, z_{i})$ in the TCS (Figure \ref{fig:systems}c).
Finally, it is important to mention that gravimeters do not measure $d_{i}$ but
$\| \mathbf{g}_i \|$.
